Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-23 draft_heyuan.Rmd
Add files via upload

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(doParallel)
library(pomp)
```

```{r data_table, echo=FALSE}
df <- read_excel("./data/bird_data.xlsx")
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

@@

Looking at both *CPUE* and *log.CPUE* in [figure], it's evident that ptarmigan populations has steadily decreased over the 142 year-long period. The blue shaded regions in plot **C** highlight years when peak rodent year is equal to "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis. It's also worth noting that this trend is far less apparent beyond the early 1900s.

Figure [] depicts the ACF plot of *log.CPUE* at 100 lags. The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values. More specifically, the initial 40 lags show strong correlation with the strongest being the first 5. The PACF plot in figure [], the partial correlation dies out after 5 lags. For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.

Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data. Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods. The repeated short-term cycle in 5 lags could be explained by the partial correlation.

@@

```{r PACF, echo=FALSE}
pacf(data$logCPUE, lag = 100, main = "PACF log.CPUE")
```
## {-}
--------

@@

@@

@@

```{r, echo=FALSE, eval=FALSE}
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1))  |>
  rename(reports = "CPUE") |>
  select(c("year", "reports", "R")) -> bird
```

```{r echo=FALSE, eval=FALSE}
birdMod |>
  pfilter(Np=5, params = params) -> pf

mif2(
  measSIR, 
  Nmif = 10, 
  Np = 2000,
  cooling.fraction.50 = 0.5, 
  rw.sd = rw_sd(
    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
  ),
  params = params
) -> mif_out
```

@@

```{r}
bird <- data[subset]
head(bird)
```

@@

```{r}
param_guess <- c(
  logF_0 = 1, 
  logB_0 = 1,
  a = 1,
  b = 2, 
  c = 1,
  alpha = 1, 
  Beta = 1,
  gamma = 0.5,
  sigmaF = 0.1, 
  sigmaB = 0.1, 
  logRho = 3, 
  sigma_obs = 0.1
)
```

```{r}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out
```

```{r}
freeze(
  runif_design(
    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
              sigmaF = 0, sigmaB = 0.02, logRho = 0)
    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
              sigmaF = 0, sigmaB = 0.02, logRho = 0)
  ),
  seed = 420
) -> guesses
```

@@

>>>>

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
library(foreach)
library(doFuture)
plan(multisession)
library(doParallel)
library(doRNG)
```

```{r data_table, echo=FALSE}
df <- read_excel("/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/data/bird_data.xlsx")
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

@@

Looking at both *CPUE* and *log.CPUE* plot above, it's evident that ptarmigan populations has steadily decreased over the 142 year-long period. The blue shaded regions in plot **C** highlight years when peak rodent year is equal to "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis. It's also worth noting that this trend is far less apparent beyond the early 1900s.

@@

```{r PACF, echo=FALSE}
pacf(data$logCPUE, lag = 100, main = "PACF log.CPUE")
```

Above figure depicts the ACF plot of *log.CPUE* at 100 lags. The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values. More specifically, the initial 40 lags show strong correlation with the strongest being the first 5. The PACF plot above, the partial correlation dies out after 5 lags. For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.

Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data. Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods, which indicates the non-existence of seasonal pattern. The repeated short-term cycle in 5 lags could be explained by the partial correlation.

@@

ARMA model can't be performed on the non-stationary time series data. The KPSS test is implemented for stationarity check, though ACF plot shows an abvious non-stationarity.  
```{r stationary, echo=FALSE}
# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss_result <- kpss.test(ts_cpue, null = "Level")
print(kpss_result)

# Differencing the time series once
start_year <- min(data$year)
end_year <- max(data$year)
frequency <- 1  
ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
ts_diff <- diff(ts_cpue, differences = 1)
print(kpss.test(ts_diff,null = "Level"))
```

KPSS with p-value = 0.01 suggests that the average daily hunting of birds(CPUE) in log scale is non-stationary. However, first-order difference on this time series data is stationary, proved by the p-value larger than $\alpha=0.05$ from KPSS test. Then the following ARMA(p,q) model will be run on such differenced data.

@@

The smaller AIC, the better the model to fit the data. Thus, ARMA(1,5) with the lowest AIC value of 217.08 is the best model to fit the differenced data. Or in other words, ARIMA(1,1,5) is the best model to fit average daily hunting of birds(CPUE) in log scale.

@@

@@

```{r, echo=FALSE, eval=FALSE}
data_path <- '/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/data/data.txt'
data <- read.table(data_path, header = TRUE, sep = "\t", na.strings = "NA", dec = ".", strip.white = TRUE)

data |>
  rename(R = "peak.rodent.year") |>
  rename(NAO = "NAO.May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

bird <- data[subset]
head(bird)
```

@@

Fisrt, we used a particle filter applied to a model over time and then assessed the performance. We gave a guess of parameter values with the following: $log(F_0) = 1, log(B_0) = 1, a = 1, b = 2, c = 1, \alpha = 1, \beta = 1, \gamma = 0.5, \sigma_F = 0.1,  \sigma_B = 0.1, log(\rho) = 3, \sigma_{obs} = 0.1$

```{r}
param_guess <- c(
  logF_0 = 1, 
  logB_0 = 1,
  a = 1,
  b = 2, 
  c = 1,
  alpha = 1, 
  Beta = 1,
  gamma = 0.5,
  sigmaF = 0.1, 
  sigmaB = 0.1, 
  logRho = 3, 
  sigma_obs = 0.1
)

mod |>
  pfilter(Np=50, params =param_guess) -> pf
plot(pf)
```

The effective sample size (ESS) in the second panel displays a presence of occasional spikes interspersed with predominantly low values. It suggests that only a few samples are providing meaningful information, which lead to a less robust and a less reliable estimate. The third panel shows the conditional log-likelihood over time which is a measure of how well the model with its current parameters. The log-likelihood appears to be fluctuating and showing some periods of higher values interspersed with periods of significant decline, indicating varying model performance over time.

Thus, our initial guess of parameter values is poor and we will do a local search of parameters for a more stablized model performance.

A single execution of the mif2 function applied to mod provides a snapshot of the model's behavior under a specific set of initial conditions.
```{r}
## local search using if2
## a single execution of the mif2 function applied to mod
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out

plot(mif2_out)
```

We used a foreach loop which iterates 20 times, each iteration of the loop applies the mif2 function to our model. This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as control the robustness.

```{r}
foreach(i=1:20,.combine=c,
    .options.future=list(seed=482947940)
  ) %dofuture% {
    mod |>
      mif2(
        Np=1000, Nmif=50,
        cooling.fraction.50=0.5,
        rw.sd = rw_sd(
           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
           a = 0.02, b = 0.02, c = 0.02, 
          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
        ),
        partrans = parameter_trans(
             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
              'c', 'logF_0', 'logB_0', 'logRho')),
        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs"),
        params = param_guess
      )
  } -> mifs_local

mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y")
```

```{r}
foreach(mf=mifs_local,.combine=rbind,
  .options.future=list(seed=900242057)
) %dofuture% {
   evals <- replicate(10, logLik(pfilter(mf,Np=500))) 
   ll <- logmeanexp(evals,se=TRUE) 
   mf |> coef() |> bind_rows() |>
   bind_cols(loglik=ll[1],loglik.se=ll[2]) } -> results

pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results,pch=16)
```

@@

```{r}

guesses <- freeze(
  runif_design(
    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
              sigmaF = 0, sigmaB = 0.02, logRho = 0),
    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
              sigmaF = 0, sigmaB = 0.02, logRho = 0),
  nseq=400 )
)

```

```{r}
set.seed(2062379496)

runif_design(
  lower = c(alpha = 0, Beta = 0, gamma = 0, a = 7, b = 2, c = 0.5,
              sigma_obs = 0, logF_0 = 0.4, logB_0 = 1.6,
              sigmaF = 0, sigmaB = 0.02, logRho = 0.3),
    upper = c(alpha = 0.1, Beta = 0.1, gamma = 0.1, a = 8, b = 3, c = 1,
              sigma_obs = 1, logF_0 = 0.5, logB_0 = 1.7,
              sigmaF = 0.1, sigmaB = 0.04, logRho = 0.5),
    nseq=400
) -> guesses

mf1 <- mifs_local[[1]]
fixed_params <- coef(mif2_out)

foreach(guess=iter(guesses,"row"), .combine=rbind,
      .options.future=list(seed=1270401374)
    ) %dofuture% {
      mf1 |>
        mif2(params=c(guess,fixed_params)) |>
        mif2(Nmif=1000) -> mf
      replicate(
        10,
        mf |> pfilter(Np=500) |> logLik()
      ) |>
        logmeanexp(se=TRUE) -> ll
      mf |> coef() |> bind_rows() |>
        bind_cols(loglik=ll[1],loglik.se=ll[2])
    } -> results

filter(is.finite(loglik)) -> results

```
